\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-30 chapters/preface.tex
Preface, ironed out overhangs

The past four years have been a stimulating, instructive, and intense
journey of scientific and self-discovery. What started as a vague interest
in the applications of machine learning in the biomedical domain during
my master's studies rapidly evolved into a fascination with neural
signal processing, from the moment I came into contact with ith.
This was supercharged by the very high science-fiction factor
of Brain-Computer Interfaces.
There is just something about both the brain and computers, and
their beautiful complexity-from-simplicity, that I just can't
seem to let go of.

@@

Furthermore, I would like to explicitly thank everyone else who helped
in realizing the work presented in this thesis. Juliette, your
proactive attitude and multidisciplinary perspective—and, of course,
your native command of the French language—were an immense help in
data collection. Without you, this thesis would probably be missing its
last chapter, and I hope you enjoy your well-deserved upcoming PhD
career. Working with participants with physical impairment would also
not have been possible without the medical advice of prof. Etienne
Allart from the Neurorehabilitation Unit at CHU Lille. I would also
like to thank the others involved in recruiting and interacting with
these participants: Alixe and dr. Edward Schietekatte at TRAINM Neuro
Rehab Clinics, prof. Kristl Claeys and prof. Philip Van Damme from the
Neuromuscular Reference Center at UZ Leuven, Marie and the team at
Fondation Partage et Vie, and all colleagues who helped me in data
collection and processing. Most importantly, I am grateful to everyone
who participated in our study, and I am sincerely moved by the positive
reception among those of you who could benefit from it. Of course, I
would also like to thank all students, colleagues, and friends who
submitted themselves to having their hair messed up in yet another EEG
experiment.

@@

In Lille, I've very much appreciated working with Gaël. Your presence
never failed to lighten up the mood, and I wish you all the best luck
in finishing up your own PhD. Next, I want to thank Jimmy for showing
me the way through Lille and its campus life, and for all the tea. I
also enjoyed discussing with José, Jean-Marc, and Marie-Hélène. Finally,
I'm also glad that the SIGMA team was around to keep me in touch with
my computer science roots. Even though I was not always physically
present in Lille, I've always considered all of you as very good
colleagues, and I'm glad I've had the opportunity to revive my high
school French, which was in dire need of a refresher.

In Leuven, I want to begin with a special thanks to those of you who
guided me in the beginning of my PhD: Benjamin, for introducing me to
the wonderful world of ERPs and linear decoding, and Axel and Arno for
powering through COVID with me. As for the rest of you lot—Aline,
Anahit, Aurélie, Barbara, Bob, Bob, Elvira, Eva, Li Ang, Liuyin, Mani,
Mani, Mansoureh, Qiang, Tjaša, Valentina, and Yide—you've made my days
brighter in many ways, and I could always count on you for advice,
support, or help. To Dilara, Gijs, Lunky, Reniflal, and Zoë, master
students and interns that I was fortunate enough to supervise: thank
you for your dedication, and good luck with your future adventures. I
would also like to mention the great work of the students at NeuroTech
Leuven—you never fail to impress me with wonderful ideas and excellent
execution. A final shoutout to everyone in and around the
Neurophysiology lab: Alessia, Astrid, Christophe, Inez, Jonathan,
Kenneth, Noa, Saba, Sara, Sofie, Stijn, Wouter, and all others.

>>>>

The past four years have been a stimulating, instructive, and intense
journey of scientific and self-discovery. What started as a vague interest
in the applications of machine learning in the biomedical domain during
my master's studies rapidly evolved into a fascination with neural
signal processing, from the moment I came into contact with ith.
This was supercharged by the very high science-fiction factor
of Brain-Computer Interfaces.
There is something about both the brain and computers, and
their beautiful complexity-from-simplicity, that I just can't
seem to let go of.

@@

Furthermore, I would like to explicitly thank everyone else who helped
in realizing the work presented in this thesis. Juliette, your
proactive attitude and multidisciplinary perspective were an immense help in
data collection. Without you, this thesis would probably be missing its
last chapter, and I hope you enjoy your well-deserved upcoming PhD
career. Working with participants with physical impairment would also
not have been possible without the medical advice of prof. Etienne
Allart from the Neurorehabilitation Unit at CHU Lille. I would also
like to thank the others involved in recruiting and interacting with
these participants: Alixe and dr. Edward Schietekatte at TRAINM Neuro
Rehab Clinics, prof. Kristl Claeys and prof. Philip Van Damme from the
Neuromuscular Reference Center at UZ Leuven, Marie and the team at
Fondation Partage et Vie, and all colleagues who helped me in data
collection and processing. Most importantly, I am grateful to everyone
who participated in our study, and I am sincerely moved by the positive
reception among those of you who could benefit from it. Of course, I
would also like to thank all students, colleagues, and friends who
submitted themselves to having their hair messed up in yet another EEG
experiment.

@@

In Lille, I've very much appreciated working with Gaël. Your presence
never failed to lighten up the mood, and I wish you smooth sailings
in completing your own PhD. Next, I want to thank Jimmy for showing
me the way through Lille and its campus life, and for all the tea. I
also enjoyed discussing with José, Jean-Marc, and Marie-Hélène. Finally,
I'm glad that the SIGMA team was around to keep me in touch with
my computer science roots. Even though I was not always physically
present in Lille, I've always considered all of you as very good
colleagues, and I'm glad I've had the opportunity to revive my high
school French, which was in dire need of a refresher.

In Leuven, I want to begin with a special thanks to those of you who
guided me in the beginning of my PhD: Benjamin, for introducing me to
the wonderful world of ERPs and linear decoding, and Axel and Arno for
powering through COVID with me. As for the rest of you lot—Aline,
Anahit, Aurélie, Barbara, Bob, Bob, Elvira, Eva, Li Ang, Liuyin, Mani,
Mani, Mansoureh, Qiang, Tjaša, Valentina, and Yide—you've made my days
brighter in many ways, and I could always count on you for advice,
support, or help. To Dilara, Gijs, Lunky, Reniflal, and Zoe, master
students and interns that I was fortunate enough to supervise: thank
you for your dedication, and good luck with your future adventures. I
would also like to mention the great work of the students at NeuroTech
Leuven—you never fail to impress me with wonderful ideas and excellent
execution. A final shoutout to everyone in and around the
Neurophysiology lab: Alessia, Astrid, Christophe, Inez, Jonathan,
Kenneth, Noa, Saba, Sara, Sofie, Stijn, Wouter, and all others.

### 2024-11-30 chapters/stbf_struct.tex
Second layout run
Preface, ironed out overhangs

\chapter{Kronecker-structured linear decoding}
\label{sec:stbf-struct}
\newcommand{\pv}[1]{
	%\ifdim#1pt<0.05pt
	%	\cellcolor{black!20}
	%\fi
	%\ifdim#1pt<0.001pt
	%	$< 0.001$
	%\else
	%	\ifdim#1pt=1.000pt
	%		\multicolumn{1}{c}{--}
	%	\else
	%		$#1$
	%	\fi
	%\fi
	#1
}
\epigraph{%
  I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
  With these hundred coins, I paid for the beast and received three coins in change.
  Of the three, I returned one coin each to my companions, keeping one for myself.
  Now, I owe each friend forty-and-nine coins, a total debt of
  ninety-and-eight. And with mine one remaining, the
  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
  Tell me then, kind stranger, whither hath the missing coin gone?
}{%
  Ancient Algerian riddle
}
% Kronecker-structured LDA ====================================================

@@

	\begin{figure}
		%\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
    \begin{subfigure}{\textwidth/4}
		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-0.eps}
      \caption{}
    \end{subfigure}\hfill%
    \begin{subfigure}{\textwidth/4}
		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-1.eps}
      \caption{}
    \end{subfigure}\hfill%
    \begin{subfigure}{\textwidth/4}
		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-2.eps}
      \caption{}
    \end{subfigure}\hfill%
    \begin{minipage}[b]{\textwidth/10}
      \begin{subfigure}{\textwidth}
  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-6.eps}
        \caption{}
      \end{subfigure}\vfill%
      \begin{subfigure}{\textwidth}
  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-7.eps}
        \caption{}
      \end{subfigure}\hfill%
    \end{minipage}

@@

	\subsection{Software and preprocessing}
	Data processing and classifier analysis were performed in Python using
	Scikit-Learn (version 1.0.1)~\cite{Pedregosa2011} and SciPy (version
	1.7.1)~\cite{Virtanen2020}.
	The preprocessing pipeline was implemented using the MNE-Python toolbox
	(version 0.24.0)~\cite{Gramfort2013}.
	The dataset was converted to BIDS-\ac{eeg} format~\cite{Pernet2019} and managed and
	loaded with MNE-BIDS (version 0.9)~\cite{Appelhoff2019}.
	The Riemannian classifier from \cref{sec:riemannian} was implemented using
	pyRiemann (version 0.2.7).
	Statistical tests were performed in R (version 4.1.2).

@@

	\subsubsection{Riemannian geometry classifier}
	\label{sec:riemannian}
	We opted for a Riemannian
	geometry-based classifier to compare our results.
	The Riemannian model (XDAWN+RG) uses the XDAWN spatial filter combined
	with Riemannian geometry in tangent space as implemented by Barachant et
	al.~\cite{Barachant2014a}.
	This classifier uses four XDAWN spatial filters and each epoch's empirical spatial covariance matrix.
	The target with the maximum score is the prediction of the cued target.
	XDAWN+RG was trained and validated without downsampling using epochs
	at the original sample rate of 2048 Hz.
  Preliminary experimentation revealed that best performance in low and high
  data availability setting was achieved when no shrinkage was applied to the estimation of XDAWN filters or the subsequent
  covariance matrices for this specific dataset.
  Hence, no shrinkage regularization will be applied to XDAWN+RG in
  further experiments.

@@

	Second, structured estimation has better time complexity.
	Covariance estimation and inversion occupy the largest part of the \ac{stbf} training time.
	For \ac{stbf-emp} and \ac{stbf-shrunk}, the time complexity of this process is $\mathcal{O}(NC^2S^2+C^3S^3)$.
	Thanks to Property~\ref{prop:stbf-struct/inverse-kronecker}, the complexity can be reduced to
	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \ac{stbf-struct}.
	The results presented in \cref{fig:stbf-struct/training-time} confirm that
  \ac{stbf-struct} is much faster in operation than other classifiers.

@@

\clearpage
\begin{subappendices}
  \section{Accuracies per number of trials and available training data}

  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
      \caption[Accuracies for \acs{stbf-struct}.]{Accuracies (\%) for \ac{stbf-struct}.}
      \label{tab:stbf-struct/accuracies/stbf-struct}
  \end{table}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
      \caption[Accuracies for \acs{stbf-shrunk}.]{Accuracies (\%) for
      \ac{stbf-shrunk}.}
      \label{tab:stbf-struct/accuracies/stbf-shrunk}
  \end{table}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
      \caption[Accuracies for \acs{stbf-emp}.]{Accuracies (\%) for
      \ac{stbf-emp}.}
      \label{tab:stbf-struct/accuracies/stbf-emp}
  \end{table}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
      \caption[Accuracies for XDAWN+RG.]{Accuracies (\%) for
      XDAWN+RG.}
      \label{tab:stbf-struct/accuracies/xdawn-rg}
  \end{table}

>>>>

\chapter{Kronecker-structured linear decoding}
\label{sec:stbf-struct}
\newcommand{\pv}[1]{
	%\ifdim#1pt<0.05pt
	%	\cellcolor{black!20}
	%\fi
	%\ifdim#1pt<0.001pt
	%	$< 0.001$
	%\else
	%	\ifdim#1pt=1.000pt
	%		\multicolumn{1}{c}{--}
	%	\else
	%		$#1$
	%	\fi
	%\fi
	#1
}
\epigraph{%
  \justifying
  I journeyed to the market to purchase a cow, its price set at ninety-and-seven coins.
  Lacking this sum, I therefore borrowed fifty coins from each of two good friends.
  With these hundred coins, I paid for the beast and received three coins in change.
  Of the three, I returned one coin each to my companions, keeping one for myself.
  Now, I owe each friend forty-and-nine coins, a total debt of
  ninety-and-eight. And with mine one remaining, the
  total merely equals ninety-and-nine coins, short of the hundred I borrowed!
  Tell me then, kind stranger, whither hath the missing coin gone?
}{%
  Ancient Algerian riddle
}
% Kronecker-structured LDA ====================================================

@@

  \begin{figure}[t]
		%\includegraphics[width=\linewidth]{figures/stbf_struct/covs.eps}
    \begin{subfigure}{\textwidth/4}
		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-0.eps}
      \caption{}
    \end{subfigure}\hfill%
    \begin{subfigure}{\textwidth/4}
		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-1.eps}
      \caption{}
    \end{subfigure}\hfill%
    \begin{subfigure}{\textwidth/4}
		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-2.eps}
      \caption{}
    \end{subfigure}\hfill%
    \begin{minipage}[b]{\textwidth/10}
      \begin{subfigure}{\textwidth}
  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-6.eps}
        \caption{}
      \end{subfigure}\vfill%
      \begin{subfigure}{\textwidth}
  		  \includegraphics[width=\linewidth]{figures/stbf_struct/covs-7.eps}
        \caption{}
      \end{subfigure}\hfill%
    \end{minipage}

@@

	\subsection{Software and preprocessing}
	Data processing and classifier analysis were performed in Python using
	Scikit-Learn (v1.0.1)~\cite{Pedregosa2011} and SciPy (version
	1.7.1)~\cite{Virtanen2020}.
	The preprocessing pipeline was implemented using the MNE-Python toolbox
	(v0.24.0)~\cite{Gramfort2013}.
	The dataset was converted to BIDS-\ac{eeg} format~\cite{Pernet2019} and managed and
	loaded with MNE-BIDS (version 0.9)~\cite{Appelhoff2019}.
	The Riemannian classifier from \cref{sec:riemannian} was implemented using
	pyRiemann (v0.2.7).
	Statistical tests were performed in R (v4.1.2).

@@

	\subsubsection{Riemannian geometry classifier}
	\label{sec:riemannian}
	We opted for a Riemannian
	geometry-based classifier to compare our results.
	The Riemannian model (XDAWN+RG) uses the XDAWN spatial filter combined
  with Riemannian geometry in tangent space as implemented by
  \textcite{Barachant2014a}.
	This classifier uses four XDAWN spatial filters and each epoch's empirical spatial covariance matrix.
	The target with the maximum score is the prediction of the cued target.
	XDAWN+RG was trained and validated without downsampling using epochs
	at the original sample rate of 2048 Hz.
  Preliminary experimentation revealed that best performance in low and high
  data availability setting was achieved when no shrinkage was applied to the estimation of XDAWN filters or the subsequent
  covariance matrices for this specific dataset.
  Hence, no shrinkage regularization will be applied to XDAWN+RG in
  further experiments.

@@

	Second, structured estimation has better time complexity.
	Covariance estimation and inversion occupy the largest part of the \ac{stbf} training time.
	For \ac{stbf-emp} and \ac{stbf-shrunk}, the time complexity of this process
  is described by $\mathcal{O}(NC^2S^2+C^3S^3)$.
	Thanks to Property~\ref{prop:stbf-struct/inverse-kronecker}, the complexity can be reduced to
	$\mathcal{O}(NS^2S^2+C^3+S^3)$ for the structured estimator of \ac{stbf-struct}.
	The results presented in \cref{fig:stbf-struct/training-time} confirm that
  \ac{stbf-struct} is much faster in operation than other classifiers.

@@

\clearpage
\begin{subappendices}
  \section{\Acs{stbf-struct} selection accuracies}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-stbf_kron_toep_shrunk.tex}
      \caption[Accuracies for \acs{stbf-struct}.]{Accuracies (\%) for \ac{stbf-struct}.}
      \label{tab:stbf-struct/accuracies/stbf-struct}
  \end{table}
  \clearpage
  \section{\Acs{stbf-shrunk} selection accuracies}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-stbf_loocv.tex}
      \caption[Accuracies for \acs{stbf-shrunk}.]{Accuracies (\%) for
      \ac{stbf-shrunk}.}
      \label{tab:stbf-struct/accuracies/stbf-shrunk}
  \end{table}
  \clearpage
  \section{\Acs{stbf-emp} selection accuracies}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-stbf_emp.tex}
      \caption[Accuracies for \acs{stbf-emp}.]{Accuracies (\%) for
      \ac{stbf-emp}.}
      \label{tab:stbf-struct/accuracies/stbf-emp}
  \end{table}
  \clearpage
  \section{XDAWN+RG selection accuracies}
  \begin{table}[!ht]
      \small
      \sffamily
	    \input{chapters/stbf_struct-acc-xdawn_rg.tex}
      \caption[Accuracies for XDAWN+RG.]{Accuracies (\%) for
      XDAWN+RG.}
      \label{tab:stbf-struct/accuracies/xdawn-rg}
  \end{table}
